Some examples in extended logic counting classes

### 2016-09-16 QSO/logics.tex
15 sep

We extend QSO with the operator $[\textbf{alfp}_{\varphi(y_1,\ldots,y_k,R)}\alpha(x_1,\ldots,x_{\ell},R,\pi)](x_1,\ldots,x_{\ell})$. It is defined by a FO-formula $\varphi$ and a QFO-formula $\alpha$ which mentions a placeholder function $\pi$ with domain $\fv^{\ell}$.

@@

As an example, we will show a formula that counts the number of paths of size $n$ of a structure $\A$ with a binary relation $E$. First we define $\varphi(x,R)$:
$$
\varphi(x,R) = \forall y(x < y \vee x = y) \vee \exists z(R(z) \wedge \varphi_{succ}(z,x)),x
$$
where $\varphi_{succ}(x,y)$ is a formula that is satisfied by pairs $(x,y)$ that are consecutive in the order $<$. That is, $\varphi_{succ}(x,y) = x < y \wedge \forall z((x < z \wedge z < y) \to (x = z \vee z = y) )$. Now we define $\alpha(R)$ as:
$$
\alpha(x,y,R) = \neg \exists z(R(z)) + \Sigma z[E(x,z)\cdot \pi(z,y)].
$$
Then, the formula $[\alfp_{\varphi(x,R)} \alpha(x,y,R)](x,y)$ when evaluated on $(\A,a,b)$ will count the number of paths of size $n$ from $a$ to $b$.

We also define the less powerful operator $\pth$ as follows. Given a positive formula $\varphi(x,R)$ and a formula $\psi(x,y)$ we define $[\pth_{\varphi(x,R)} \psi(x,y)](x,y)$ as a formula that when evaluated on $(\A, a, b)$ induces a graph over $\A$ with the binary relation determined by $\psi(x,y)$ and counts the number of paths from $a$ to $b$ of a size determined by the $\varphi(x,R)$. Formally, we define:
$$
\sem{[\pth_{\varphi(x,R)} \psi(x,y)](x,y)} = \sem{[\alfp_{\varphi(x,R)} \alpha_{\psi}(x,y,R)](x,y)},
$$
where $\alpha_{\psi}(x,y,R) = \neg \exists z(R(z)) + \Sigma z[\psi(x,z)\cdot \pi(z,y)]$.

>>>>

We extend $\qfo$ with the operator $[\textbf{alfp}_{\varphi(y_1,\ldots,y_k,R)}\alpha(x_1,\ldots,x_{\ell},R,\pi)](x_1,\ldots,x_{\ell})$. It is defined by a FO-formula $\varphi$ and a QFO-formula $\alpha$ which mentions a placeholder function $\pi$ with domain $\fv^{\ell}$.

@@

As an example, we will show a formula that counts the number of paths of size $n$ of a structure $\A$ with a binary relation $E$. First we define $\varphi(x,R)$:
$$
\varphi(x,R) = \forall y(x < y \vee x = y) \vee \exists z(R(z) \wedge \varphi_{succ}(z,x)),x
$$
where $\varphi_{succ}(x,y)$ is a formula that is satisfied by pairs $(x,y)$ that are consecutive in the order $<$. That is, $\varphi_{succ}(x,y) = x < y \wedge \forall z((x < z \wedge z < y) \to (x = z \vee z = y) )$. Now we define $\alpha(x,y,R,\pi)$ as:
$$
\alpha(x,y,R,\pi) = \neg \exists z(R(z)) + \Sigma z[E(x,z)\cdot \pi(z,y)].
$$
Then, the formula $[\alfp_{\varphi(x,R)} \alpha(x,y,R,\pi)](x,y)$ when evaluated on $(\A,a,b)$ will count the number of paths of size $n$ from $a$ to $b$.

We also define the less powerful operator $\pth$ as follows. Given a positive formula $\varphi(x,R)$ and a formula $\psi(x,y)$ we define $[\pth_{\varphi(x,R)} \psi(x,y)](x,y)$ as a formula that when evaluated on $(\A, a, b)$ induces a graph over $\A$ with the binary relation determined by $\psi(x,y)$ and counts the number of paths from $a$ to $b$ of a size determined by the $\varphi(x,R)$. Formally, we define:
$$
\sem{[\pth_{\varphi(x,R)} \psi(x,y)](x,y)} = \sem{[\alfp_{\varphi(x,R)} \alpha_{\psi}(x,y,R,\pi)](x,y)},
$$
where $\alpha_{\psi}(x,y,R,\pi) = \neg \exists z(R(z)) + \Sigma z[\psi(x,z)\cdot \pi(z,y)]$.
